---
title: "Adding Census Population to ACFRs Data at City and Town Level"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
options(scipen = 999)
library(tidyverse)
library(dplyr)
library(DT)
```
# Tasks 

*City names in Acfrs database are not obvious, i.e., does not contain a certain word to indicate it's a city

*The [census population data use GEOIDs](https://www.census.gov/programs-surveys/geography/guidance/geo-identifiers.html#:~:text=The%20full%20GEOID%20for%20many,codes%2C%20in%20which%20they%20nest.), which are "numeric codes that uniquely identify all administrative/legal and statistical geographic areas for which the Census Bureau tabulates data". 


*ACFRs data has government_id (which is called census_id in the ACFRs portal), but not geo_id. 

*Use a "middle file" to link geo_id and government_ID

*Need to a) identify city and towns in census population data and b) link those to corresponding acfrs entities. 

# Census Polulation - Cities and Towns

List of cities and towns with geo_id. 

```{r}
#Marc emailed this data Feb2 2022: using 017 Census master file data to map to the 2020 Census Population data using FIPS Codes. GEO_ID: Census identifiers that point to population of places in US
geoID_pop20_census <- rio::import(here::here("data", "City and Town Mapping.xlsx"), sheet = 2) %>% 
rename(population = `!!Total:`, 
       geo_id = GEO_ID) 

# join with df_state to get state.abb and state.name --> to have a common col with acfrs to join later
df_state <- data.frame(state.abb, state.name) %>% 
  add_row(state.abb = "DC", state.name = "District of Columbia")


pop <- geoID_pop20_census %>% 
  separate(NAME, c("city_town", "state.name"), sep = ",")  %>%  
  mutate(state.name = str_trim(state.name), 
         city_town = str_trim(city_town)) %>% 
  left_join(df_state) %>% drop_na()

datatable(pop)

```

There are `r nrow(pop)` observations in this data set. 

## Linking government_ID and geo_id

ACFRs database uses government ID (enumeration ID of local governments), but do not have geo_id. 

We use a Census master file data to map ACFRs to the 2020 Census Population data by FIPS Codes.

```{r}
#government_ID (used in ACFRs): enumeration ID of local governments. ACFRs data does not use FIP codes 
# Note that population data is ACFRs portal is 2017 -> don't use
# Note: census_id, which means government_id in Acfrs are numeric --> can be turn to scientific notation in R --> will not match --> need to avoid scientific notation. 

# This has both government_ID and geo_id
governmentID <- rio::import(here::here("data", "City and Town Mapping.xlsx"), sheet = 3) %>% 
  select(government_ID, `INFERRED GEO_ID`, NAME, CITY, STATE_AB, COUNTY_AREA_NAME)  %>% 
  rename(geo_id = `INFERRED GEO_ID`,  # Marc created INFERRED GEO_ID, which meant to be geo_id
         original_name = NAME)   
  
# Use governmentID to join with population
pop_governmentID <- pop %>% 
  left_join(governmentID) %>% drop_na(government_ID)

datatable(pop_governmentID %>% select(geo_id, government_ID, city_town, population))
```

# City and Towns

*Get all general purpose entities in acfrs 

*Join above acfrs entities with dataset of city and town population data that has governmentID

```{r}
# census_id in ACFRs is government_ID used in file "City and Town Mapping.xlsx"

# 1. get all general purpose entities in acfrs
acfrs_governmentID <- readRDS("data/data_from_dbsite_2020.RDS") %>% 
  filter(category == "General Purpose") %>% 
  rename(government_ID = census_id) %>% 

# problem: government_ID in pop_governmentID has 14 characters (05201902700000) >< some government_ID in ACFRs has 13 characters 5201902700000 --> Marc: Yes, it is safe to assume that these government_ID in ACFRs data should have 0 at the beginning.
  mutate(government_ID = ifelse(str_length(government_ID) < 14, paste0("0", government_ID), government_ID)) 


# 2. join with city and town population data that has governmentID
acfrs_city_pop <- acfrs_governmentID %>% 
  left_join(pop_governmentID) %>% 
  select(-c(state, STATE_AB, year)) %>% 
  arrange(desc(population)) %>% 
  drop_na(population)

datatable(acfrs_city_pop)
```
At this point Acfrs now has `r nrow(acfrs_governmentID)` general purpose entities FY 2020, of which 6810 are cities and towns. 

## Checking for overlapping 

Check if there's any overlapping between the above lists

```{r}
# all general purpose entities: 11187
length(unique(acfrs_governmentID$government_ID)) 

unique(acfrs_governmentID$government_ID) -> t1

# cities and towns: 6810
length(unique(acfrs_city_pop$government_ID)) 
unique(acfrs_city_pop$government_ID) ->t2

# gap bw the 2 above: 4377
length(unique(setdiff(t1, t2)))

# list of counties (in count_gov.Rmd): 2569
county_pop_census_acfrs <- readRDS("county_pop_census_acfrs.RDS")
length(unique(county_pop_census_acfrs$census_id))
unique(county_pop_census_acfrs$census_id) ->t3

# entities overlapping  between city list and county list

intersect(t2, t3) -> intersect_city_county

# view the overlapping cities 
acfrs_city_pop %>% filter(government_ID %in% intersect_city_county) %>% 
 select(name, government_ID)

```
Those whose names contain parentheses are actually cities. The part in side the parentheses indicates which county they belong to. 


## Final City and County List

```{r}
state_gov <-readRDS("state_gov.RDS")

# composition of city, county, state gov in general purpose
 # acfrs_governmentID %>% 
 #  left_join(pop_governmentID) %>% #drop_na(population) %>% 
 #  select(-c(state, STATE_AB, year)) %>% 
 #  arrange(desc(population)) %>% 

# take out county 2569
# filter(!government_ID %in% county_pop_census_acfrs$census_id) %>% 
#    filter(!id %in% state_gov$id)
   
```



Lastly, take 15 who are actually counties out of the cities list. 
```{r}
# actually county
  acfrs_city_pop %>% filter(government_ID %in% intersect_city_county) %>% 
  filter(!grepl("\\)", name)) -> actual_county

acfrs_city_pop_final <- acfrs_city_pop %>% 
  filter(!government_ID %in% actual_county$government_ID) 

datatable(acfrs_city_pop_final)
 saveRDS(acfrs_city_pop_final, "acfrs_city_pop.RDS")
```

The list of city and county contains `r nrow(acfrs_city_pop_final)` entities. 

Another approach: 

Subtract all states and counties from general purpose

```{r}
state_gov <- readRDS("state_gov.RDS")
counties <- readRDS("counties.RDS") 

gen_purpose <- readRDS("data/data_from_dbsite_2020.RDS") %>% 
  filter(category == "General Purpose")  
  
gen_purpose %>% 
  mutate(id = as.character(id)) %>% 
  # take out state gov
  filter(!id %in% state_gov$id) %>% 
  
  # take out county %>% 
  filter(!id %in% counties$id.x)
  

  
  # check id in city and county

length(setdiff(acfrs_city_pop_final$id, counties$id.x))

#over lapping city and county
length(intersect(acfrs_city_pop_final$id, counties$id))
  
```

```{r}
nrow(counties)
slices <- c(51, nrow(counties), nrow(gen_purpose) - nrow(counties))
lbls <- c("state", "county", "city and township")

lbls <- paste(lbls,":", slices, "entities")
pie(slices, labels = lbls, main="Share of entities in General Purpose")

```


```{r}
slices <- c(51, nrow(counties), nrow(acfrs_city_pop_final), nrow(gen_purpose) - nrow(counties)-nrow(acfrs_city_pop_final))
lbls <- c("state", "county", "city and township \nwith pop data", "city and township \nw/o pop data")

lbls <- paste(lbls,":", slices, "entities")
pie(slices, labels = lbls, main="Share of entities in General Purpose with \nand w/o population data")
```

```{r}
# general purpose w/o pop data:
gen_purpose %>% 
  filter(!id %in% state_gov$id) %>% 
  filter(!id %in% counties$id.x) %>% 
  filter(!id %in% acfrs_city_pop_final$id) %>% 
  
  # 
   # filter(total_liabilities > 500000000) %>% # 33 entities is 5 hundred mil a resonable cut? Using total or using net_pension
   # arrange(desc(total_liabilities))

# Using total or using net_pension
  arrange(desc(net_pension_liability)) %>% 
  filter(net_pension_liability > 100000000) # 42


# using both: 

  filter(total_liabilities > 500000000 & net_pension_liability > 100000000) # 23 total_lia more than 500 mil & net pen sion more than 100 mil
  

# Manually adding population data to entities with total_liability greater than 500 mil

  
  
  
```

