---
title: "Matching Census Population to ACFRs Data at County Level"
author: "TN"
date: January 28, 2022
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(dplyr)

```

# Get County Population from Census 

```{r}
d_county <- rio::import(here::here("data", "DECENNIALPL2020.P1_data_with_overlays_2021-12-16T123049.csv"), skip = 1) %>% 
  select(id, `Geographic Area Name`, `!!Total:`) %>% 
  rename(population = `!!Total:`)

# join with df_state to get state.abb and state.name --> to have a common col to join with acfrs
df_state <- data.frame(state.abb, state.name) %>% 
  add_row(state.abb = "PR", state.name = "Puerto Rico") %>% 
  add_row(state.abb = "DC", state.name = "District of Columbia")

pop_county <- d_county %>% 
  separate(`Geographic Area Name`, c("county", "state.name"), sep = ",")  %>%  
  mutate(state.name = str_trim(state.name)) %>% 
  left_join(df_state) %>% 
  
# normalize county names in pop
  mutate(county = str_to_lower(county),
         county = str_remove(county,"\\.|'|‘")) %>% 

# Nashville-Davidson metropolitan government (balance) TN == "Nashville and Davidson County" TN in ACFRS--> rename to match
mutate(county = ifelse(county == "nashville-davidson metropolitan government (balance)" & state.abb == "TN", "nashville and davidson county", county)) %>%
# 
# # Louisville/Jefferson County metro government (balance) KY == Louisville-Jefferson County KY in ACFRs 
mutate(county = ifelse(county == "louisville/jefferson county metro government (balance)" & state.abb == "KY", "louisville-jefferson county", county))


## Check some special cases: Names in Census population differ from names in ACFRs
  # pop_county %>% 
  # filter(state.abb == "TN" | state.abb == "KY") %>% 
  # filter(str_detect(county, "davidson|jefferson")) #jefferson county ; davidson county
# 
# # change to name style in acfrs
# mutate(county = ifelse(county == "kauai county", "kaua‘i county", county))
# 
# # recheck special cases: 
# pop %>% 
#   filter(str_detect(county, "nashville-davidson|louisville-jefferson|kauai county"))
```

There are `r nrow(pop_county)` observations in census population data.

# ACFRs data - County Level

```{r}
acfrs <- readRDS("data/data_from_dbsite_2020.RDS") %>% 
# rename to easily join with census pop data later
  rename(county = name) %>% 
  rename(state.abb = state) %>% 
  
# normalize county names 
  mutate(county = str_to_lower(county), 
         county = str_remove(county,"\\.|'|‘")
         ) #%>% filter(county == "nashville and davidson county")
```
ACFRs has `r nrow(acfrs)` observations. 


## Find "County" or "Municipality" in ACFRs
```{r}
# ACFRs entities that contains the word "County" in their names
  acfrs_county <- acfrs %>% 
  filter(category == "General Purpose") %>% 
  filter(grepl("county|municipality", county)) 

```

There are `r nrow(acfrs_county)` entities that contain the word "county" in their names. 

## Special cases 
### Louisiana

In Louisiana, counties are called Parishes.

```{r}
louisiana_parish <- acfrs %>% 
  filter(category == "General Purpose") %>%  
  filter(state.abb == "LA") %>% 
  filter(grepl("parish", county))

```
There're `r nrow(louisiana_parish)` Louisiana entities that contain the word "Parish" in their names. 

### Alaska

Alaska often uses the term "Borough" instead of County. 
```{r}
#Check term “Municipality” too, not found any. 
alaska_borough <- acfrs %>% 
  filter(category == "General Purpose") %>%  
  filter(state.abb == "AK") %>% 
  filter(grepl("borough", county)) %>% arrange(county)
          
```

### Puerto Rico (Note: not using to match with ACFRS for now)
Update Feb 7/2022: Marc: "Drop Puerto Rico out of the county analysis?  I realize it
was in the 2020 Census data, but they are not going to be on our map."

Puerto Rico only has “Municipios” that perform the functions of both cities and counties.

The problem is: 
*In ACFRs, there's no Puerto Rico entities that has "Municipio" in their names.
*In Census, Puerto Rico counties do contain the word "Municipio" in their names.

Solution: Remove "Municipio" in the names in Census to match with ACFRs

Potential risk: Not all Puerto Rico entities in ACFRs without the word "Municipio" are actually a municipio.

```{r}
# in Census population, find county that has the word "Municipio" --> remove the word "Municipio" to match with acfrs 
puertorico_census_pop <- pop_county %>% 
          filter(state.name == "Puerto Rico") %>% 
  filter(grepl("municipio", county)) %>% 
  mutate(county = str_remove(county, " municipio")) 
```

```{r}
# In ACFRs, get all PR entities in General Purpose 
 puertorico_afrs_census_pop <-  acfrs %>% 
          filter(category == "General Purpose") %>% 
          filter(state.abb == "PR") %>% 
          #join Puerto Rico in acfrs with PR in census -
  left_join(puertorico_census_pop, by = c("state.abb", "county")) 
```

# Join ACFRs and Census County Population 

Joining these components: 

*acfrs entities contain word "County" or "Municipality" in their names

*acfrs entities of Louisiana that contain word "Parish" in their names

*acfrs entities of Louisiana that contain word "Borough" in their names

*Puerto Rico in acfrs and Census. Note that in ACFRs, there's no Puerto Rico entities has "Municipio". To match with PR in Census, need to remove this word from Census population data. This introduces a risk that there might be cases where PR entities in ACFRs without word "Municipio" are NOT actually Municipio.


```{r}
# first, join entities in ACFRs contain words "County" + Louisiana that contain word "Parish" + alaska that contain "Borough" in their names
acfrs_county_parish_borough <- rbind(acfrs_county, louisiana_parish) %>% 
                              rbind(alaska_borough) 

# next, join with census Census population by = c("state.abb", "county")
county_pop_census_acfrs <- acfrs_county_parish_borough %>% 
            left_join(pop_county, by = c("state.abb", "county")) %>% 
            
# third, bind with puertorico 
           #rbind(puertorico_afrs_census_pop) %>% 
            arrange(desc(population)) %>% distinct() 

# --> 2569 county level entities, of which 179 do not have population data
#county_pop_census_acfrs %>% filter(is.na(population)) 

```


```{r}
#double check davidson|jefferson
county_pop_census_acfrs %>% 
  filter(state.abb == "KY" | state.abb == "TN") %>% 
  filter(str_detect(county, "davidson|jefferson")) %>% select(state.abb, population, total_liabilities, id.y, county, census_id)
  
#write.csv(county_pop_census_acfrs, "county_pop_census_acfrs.csv")
saveRDS(county_pop_census_acfrs, "county_pop_census_acfrs.RDS")
```

The matched dataset has `r nrow(county_pop_census_acfrs)` observations/ counties. 

# County Data Analysis 

## Calculate Ratio, Per Capita

```{r}
# To map
counties_1 <- county_pop_census_acfrs %>% 
  mutate(lib_rev_ratio = (total_liabilities/revenues)*100,
         lib_per_capita = total_liabilities/population, 
         county = str_to_title(county)) %>% 
  
  #rename cols to match Jordan's original files to display on map
  rename(FIPS = id.y, 
         name = county) %>%
  
  mutate(FIPS = str_sub(FIPS, -5, -1)) %>% 
  arrange(desc(population)) %>% 
  select(FIPS, state.name, name, population, total_liabilities, revenues, lib_rev_ratio, lib_per_capita, state.abb) %>% arrange(desc(population)) 

```

## Special cases: Combined city/ county government

Some cities are counties. Need to extract the data of cities list to add new rows in counties list. 

*Jacksonville city, FL  = Duval County, FL

*San Francisco city, CA = San Francisco County, CA

*"City and County of Denver" = "Denver County"

*"Philadelphia" = "Philadelphia County"

*"Indianapolis" = "Indianapolis-Marion County"

```{r}

acfrs_city_pop <- readRDS("acfrs_city_pop.RDS") # produced from city_gov.Rmd

##Jacksonville city, FL  = Duval County, FL 
# copy the data for Jacksonville city, FL to a new row in the County data for Duval County, FL
jacksonville_duval <- acfrs_city_pop %>% 
filter(state.abb == "FL" & name == "Jacksonville") %>% 
  # adding FIPS code manually
  mutate(FIPS = "12031", # got it here https://www.census.gov/quickfacts/fact/table/duvalcountyflorida,US/PST045221
      name = "Duval County")
  
## San Francisco city, CA = San Francisco County, CA
sanfrancisco <- acfrs_city_pop %>% 
filter(state.abb == "CA" & name == "San Francisco") %>% 
  mutate(FIPS = "06075",
         name = "San Francisco County")

#Denver, CO
denver <- acfrs_city_pop  %>% 
filter(state.abb == "CO" & name == "City and County of Denver") %>%
  mutate(FIPS = "08031",
         name = "Denver County")


#Philadelphia, PA
philadelphia <- acfrs_city_pop %>% 
filter(state.abb == "PA" & name == "Philadelphia") %>% 
  mutate(FIPS = "42101", 
         name =  "Philadelphia County")
         

#Indianapolis-Marion County, Indiana
indianapolis_marion <- acfrs_city_pop %>% 
filter(state.abb == "IN" & name == "Indianapolis") %>% 
  mutate(FIPS = "1836003", # Indianapolis city (balance), Indiana
         name = "Indianapolis-Marion County")
```


```{r}
# Rbind the special cases
counties <- rbind(jacksonville_duval, 
                  sanfrancisco,
                  denver,
                  philadelphia,
                  indianapolis_marion) %>% 
  mutate(lib_rev_ratio = (total_liabilities/revenues)*100,
         lib_per_capita = total_liabilities/population) %>% 
  select(FIPS, state.name, name, population, total_liabilities, revenues, lib_rev_ratio, lib_per_capita, state.abb) %>% 

# rbind to the rest of counties
  rbind(counties_1)
  
  
#re check all special cases: 
counties  %>% 
  filter(name == "Duval County" |
                      name == "San Francisco County" |
                      #name == "Denver County" |
                      name == "Philadelphia County" |
                      name == "Indianapolis-Marion County" |
                      name == "Louisville-Jefferson County" |
                      name == "Nashville-Davidson County"
                      ) %>% 
  filter(state.name != "Texas") # not Duval in TX

# check other cases if they all have pop data: Kaua‘i County, HI; Dekalb County, AL; Desoto County, FL; St Francis County, AR
counties %>% 
  filter(str_detect(name, "Kauai|Dekalb|Desoto|Francis"))

saveRDS(counties, "counties.RDS")
write.csv(counties, "counties.csv")
```