---
title: "Untitled"
output: html_document
date: "2023-02-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 The below use file cities.csv --> not sustainable --> redo above
```{r}
# Some cities are counties --> get from list of 100 cities
cities <- read.csv("cities.csv") %>% 
select(-c(lat, long, X)) %>% rename(name = city, # to match with colnames in counties file
                                 revenue = revenues) 

##Jacksonville city, FL  = Duval County, FL 
# copy the data for Jacksonville city, FL to a new row in the County data for Duval County, FL

jacksonville_duval <- cities %>% 
filter(state_abv == "FL" & name == "Jacksonville") %>% 
mutate(percapita = total_liabilities/population,
 # adding FIPS code manually
      FIPS = "12031", # got it here https://www.census.gov/quickfacts/fact/table/duvalcountyflorida,US/PST045221
      name = ifelse(name == "Jacksonville", "Duval County", name))

## San Francisco city, CA = San Francisco County, CA
sanfrancisco <- cities %>% 
filter(state_abv == "CA" & name == "San Francisco") %>% 
  mutate(percapita = total_liabilities/population, 
         FIPS = "06075",
         name = ifelse(name == "San Francisco", "San Francisco County", name))

```


```{r}
#Denver, CO
denver <- cities  %>% 
filter(state_abv == "CO" & name == "Denver") %>% 
  mutate(percapita = total_liabilities/population, 
         FIPS = "08031", 
         name = ifelse(name == "Denver", "Denver County", name)) 

```


```{r}
#Philadelphia, PA
philadelphia <- cities %>% 
filter(state_abv == "PA" & name == "Philadelphia") %>% 
  mutate(percapita = total_liabilities/population, 
         FIPS = "42101", 
         name = ifelse(name == "Philadelphia", "Philadelphia County", name))
```

```{r}
#Indianapolis-Marion County, Indiana
indianapolis_marion <- cities %>% 
filter(state_abv == "IN" & name == "Indianapolis") %>% 
  mutate(percapita = total_liabilities/population, 
         FIPS = "1836003", # Indianapolis city (balance), Indiana
         name = ifelse(name == "Indianapolis", "Indianapolis-Marion County", ""))
```



# List of counties in Census data with > 100k population  

```{r}
# stoplist

# stoplist1 <-  pop %>% 
#   filter(state.name == "New York" & county %in% c("Bronx County", "Richmond County", "Kings County", "Queens County", "New York County"))
#   
# stoplist2 <- pop %>% 
#   filter(state.name == "Florida" & county == "Duval County") # FL Duval County's government is combined with FL Jacksonville,
# 
# stoplist3 <- pop %>% 
#     filter(state.name == "Tennessee" & county == "Davidson County") 
#     
# stoplist4 <-  pop %>% 
#   filter(state.name == "Kentucky" & county == "Jefferson County")
# 
# stoplist5 <- pop %>% 
#     filter(state.name == "Ohio" & county == "Franklin County")
#   
# stoplist <- rbind(stoplist1, stoplist2, stoplist3, stoplist4, stoplist5)
# 
# # get counties in Census population data that have > 100k pop, county level, not Connecticut, not in stoplist
# 
# census_pop_100k <- pop %>% 
#   # counties with > 100k  pop in Census
#   filter(population > 100000) %>% 
#   
#   # only get county level 
#   filter(grepl("County|Parish", county)) %>%  # how about |Borough?
#   filter(state.name != "Connecticut") %>% 
#   filter(!id %in% stoplist$id) 

```

## Counties in Census with > 100k population above that are not yet matched with ACFRs

```{r}
# census_pop_NOT_match_acfrs_100k <- census_pop_100k %>% 
#   filter(!id %in% county_pop_census_acfrs$id ) %>% arrange(desc(population)) %>% distinct()
#write.csv(census_pop_NOT_match_acfrs_100k, "census_pop_NOT_match_acfrs_100k.csv")
```

There are `r nrow(census_pop_NOT_match_acfrs_100k)` counties in Census with more than 100k population that are not yet matched with ACFRs. 

## All counties in Census, regardless of population size, that are not yet matched with ACFRs

county_pop_census_acfrs

```{r}
# census_pop_NOT_match_acfrs_all <- pop %>% 
# 
#   # only get county level 
#   filter(grepl("County|Parish", county)) %>%  # how about |Borough?
#   filter(state.name != "Connecticut") %>%
#   filter(!id %in% stoplist$id) %>% 
#   filter(!id %in% county_pop_census_acfrs$id) %>% arrange(desc(population))

#write.csv(census_pop_NOT_match_acfrs_all, "census_pop_NOT_match_acfrs_all.csv")
```

There are `r nrow(census_pop_NOT_match_acfrs_all)` counties in Census, regardless of population size, that are not yet matched with ACFRs. 

```{r}
#++++++++
#Note: the below match is NOT accurate: it will filter out counties whose names appear in county_pop_census_acfrs. 
#Some states have counties with the same name. 
# Eg. this code will filter out "Kings County", both CA and NY. 
# WHile Kings County NY is not in ACFRs, Kings County CA is. 

#  filter(!county %in% county_pop_census_acfrs$county) %>% distinct()
#+++++++++
```


